\chapter{The Condensed Checklist}
\label{app:checklist}

The whole book as an operational sequence. Each item cites its chapter.

\section*{Before modeling}
\begin{enumerate}[label=\arabic*., itemsep=1pt]
\item Locate the metric's zero point and the realistic ceiling; write both
  at the top of your notes. (Ch.~\ref{ch:problem}, \ref{ch:metric})
\item Read the metric's formula character by character; reimplement it
  yourself; push weights into every loss and every early-stopping
  criterion. (Ch.~\ref{ch:metric})
\item Fingerprint every target: ACF shape (ramps? cutoffs?), spectrum
  (nulls? floors?), shift scan ($R^2$ vs.\ relative lag).
  (Ch.~\ref{ch:targets})
\item Fingerprint every feature the same way; build the lead--lag atlas
  against the (deconvolved) reference signal; split lead from nowcast.
  (Ch.~\ref{ch:featforensics})
\item Pass every structural finding through the two-window stability gate
  before it becomes load-bearing. (Ch.~\ref{ch:featforensics},
  \ref{ch:negative})
\item Write the deployment information set as one formal sentence; derive
  the forensic report and the dead-ends list. (Ch.~\ref{ch:leakage},
  \ref{ch:targets})
\end{enumerate}

\section*{Data discipline}
\begin{enumerate}[label=\arabic*., itemsep=1pt, resume]
\item Type every feature: trailing / contemporaneous / forward; forward
  types are targets only. Lags via date-shifted joins.
  (Ch.~\ref{ch:leakage})
\item Fit all preprocessing on training dates only; freeze at the
  boundary; online updates from predicted-past days only.
  (Ch.~\ref{ch:preprocessing})
\item Embargo $\ge$ the pipeline's longest forward reach --- labels
  included. (Ch.~\ref{ch:leakage})
\item Walk-forward validation that includes the adaptation procedure;
  static and online scores reported separately.
  (Ch.~\ref{ch:validation})
\item Establish CV--LB correlation with a handful of spread-out probe
  submissions before trusting either. (Ch.~\ref{ch:validation})
\item Measure your seed noise; effects inside one seed-noise unit do not
  exist yet. Two-tier validation; pre-registered experiments;
  append-only log. (Ch.~\ref{ch:problem}, \ref{ch:diagnostics})
\item Precompute features once; float16 storage / float32 compute;
  memmap with a date-range manifest; rows ordered by access pattern.
  (Ch.~\ref{ch:preprocessing})
\end{enumerate}

\section*{Features}
\begin{enumerate}[label=\arabic*., itemsep=1pt, resume]
\item Decompose, don't destroy: level + innovation; market +
  idiosyncratic; keep both parts. (Ch.~\ref{ch:smoothing},
  \ref{ch:preprocessing})
\item Never smooth a timing signal; difference or deconvolve
  pre-smoothed series at their detected window.
  (Ch.~\ref{ch:smoothing})
\item Build the cross-sectional block (market averages, ranks) --- the
  no-latency direction. (Ch.~\ref{ch:featurepatterns})
\item Convert construction knowledge into algebra: multi-scale
  differences, venue spreads, realized/nowcast auxiliary targets.
  (Ch.~\ref{ch:featurepatterns})
\item Selection: fold-consistent importance $\cup$ forensic seeds;
  symbolic search with parent-beating and redundancy caps; prune raw
  dumps, never curated pools; schedule recycling.
  (Ch.~\ref{ch:selection})
\item Ablate every block on the model family that will consume it, on
  the fixed harness. (Ch.~\ref{ch:featurepatterns})
\end{enumerate}

\section*{Models}
\begin{enumerate}[label=\arabic*., itemsep=1pt, resume]
\item XGBoost first --- baseline, meter, and permanent ensemble member.
  (Ch.~\ref{ch:treesmlp})
\item Track the XGB--MLP gap as your feature-quality dashboard.
  (Ch.~\ref{ch:treesmlp})
\item A recurrent model with auxiliary heads is the presumptive
  workhorse wherever labels arrive with delay; day = batch;
  clip gradients. (Ch.~\ref{ch:rnn})
\item Attention only as a targeted bet on forensically evidenced
  structure, made causal for the protocol, and triaged on a synthetic
  world first. (Ch.~\ref{ch:attention}, \ref{ch:synthetic})
\item Judge every candidate family by the two taxes: adaptation
  (its own refit-lr sweep) and correlation (the admission rule).
  (Ch.~\ref{ch:online}, \ref{ch:ensembling})
\end{enumerate}

\section*{Training and operations}
\begin{enumerate}[label=\arabic*., itemsep=1pt, resume]
\item Recalibrate your eyes: loss moves in the third decimal; epoch~1 is
  negative; $+0.008$ is good; $+0.05$ is a leak.
  (Ch.~\ref{ch:diagnostics})
\item Debug up the ladder --- floor anchors, deliberate overfit,
  shift/shuffle probes, seed replication --- before touching
  hyperparameters. (Ch.~\ref{ch:diagnostics})
\item Online refit on by default; sweep its learning rate per family;
  expect plateaus a decade wide and cliffs on both sides.
  (Ch.~\ref{ch:online})
\item Sweep the recency window; expect a knee; do not train on ancient
  regimes. (Ch.~\ref{ch:online}, \ref{ch:negative})
\item Ensemble: simple average of best-per-family; diversity from
  families, seeds, temporal bags, adaptation variants; climb the
  weighting ladder only while untouched data pays.
  (Ch.~\ref{ch:ensembling})
\item Bury every dead idea in the graveyard file with its autopsy;
  identify which of the four pathogens got it.
  (Ch.~\ref{ch:negative})
\end{enumerate}
