\chapter{Glossary}
\label{app:glossary}

\begin{description}[style=nextline, itemsep=2pt]

\item[ACF (autocorrelation function)] Correlation of a series with itself
at each lag. The genre's primary fingerprint reader: triangular ramp with
cutoff at $w$ $\Rightarrow$ SMA($w$) of noise; geometric decay
$\Rightarrow$ AR/level series. (Ch.~\ref{ch:targets})

\item[Alpha score] Atlas summary: a feature's mean correlation with the
(deconvolved) underlying signal over future lags inside the target's
window --- the predictive part. Contrast \emph{nowcast score}.
(Ch.~\ref{ch:featforensics})

\item[Auxiliary targets / aux heads] Additional supervised outputs
(other responders, synthetic forward or realized quantities) trained
alongside the main target to shape a shared representation; legal
because targets may look forward. (Ch.~\ref{ch:featurepatterns})

\item[Ceiling (noise ceiling)] The maximum achievable score given the
information set --- the variance share of the target that is predictable
at all. Estimate before modeling. (Ch.~\ref{ch:problem})

\item[Contemporaneous feature] Computed from data delivered at prediction
time $t$ (e.g.\ the cross-symbol market average at a timestamp). Legal
under this protocol. (Ch.~\ref{ch:leakage})

\item[Cross-sectional] Across symbols at a fixed timestamp --- the panel
direction with no temporal latency. (Ch.~\ref{ch:featurepatterns})

\item[Deconvolution] Inverting a known linear filter (e.g.\ an SMA) to
recover the underlying series; regularized (ridge/Tikhonov) because
additive noise makes exact inversion ill-posed.
(Ch.~\ref{ch:featforensics})

\item[Drift] Slow change in the feature--target relationship over time.
Causes: regime change, adaptation by market participants, pipeline
changes. Antidotes: recency windows, online refit.
(Ch.~\ref{ch:problem}, \ref{ch:online})

\item[Embargo] A buffer of discarded data between training and validation
periods, at least as wide as the pipeline's longest forward reach
(labels included). (Ch.~\ref{ch:leakage})

\item[Endogenous / exogenous split] Architectural separation of the
target's own history (fully known, non-causally attendable when
delivered as daily lags) from streaming covariates (must stay causal).
TimeXer's organizing idea. (Ch.~\ref{ch:attention})

\item[Fold-consistency (SHAP-consistency)] Feature-selection criterion:
keep features important in (nearly) all temporally distinct CV folds,
not just on average --- a stability filter on importance.
(Ch.~\ref{ch:selection})

\item[Lead--lag atlas] Per-feature curves
$\corr(x_t, \hat s_{t+k})$ over lags $k$, summarized into alpha/nowcast/
peak statistics; the campaign's forensic centerpiece.
(Ch.~\ref{ch:featforensics})

\item[Leakage] Any pathway by which information unavailable at prediction
time (temporally or by protocol) enters model inputs or preprocessing
statistics. (Ch.~\ref{ch:leakage})

\item[Medoid] Cluster representative: the member with maximal total
similarity to its clustermates; used to prune redundant raw feature
dumps. (Ch.~\ref{ch:selection})

\item[Nowcast score] Atlas summary: correlation with the just-realized
signal (backward lags) --- descriptive content. High-nowcast features
explain the present, not the future. (Ch.~\ref{ch:featforensics})

\item[Nowcast auxiliary head] An aux target equal to the
\emph{realized} (completed) version of the target's window --- a
high-SNR ($R^2 \approx 0.5$ here) training task anchoring the shared
representation. (Ch.~\ref{ch:featurepatterns})

\item[Online refit] A small gradient update on newly revealed labels
(here: daily) during the evaluation/deployment walk; the genre's
largest lever, with family-specific learning-rate plateaus and cliffs.
(Ch.~\ref{ch:online})

\item[Panel] Data indexed by (entity, date, time) --- many series
observed on a shared clock. (Ch.~\ref{ch:problem})

\item[Purged $k$-fold] Time-series CV with contiguous folds where
training rows whose label windows overlap the validation fold are
removed, plus an embargo. (Ch.~\ref{ch:validation})

\item[Recency window] Training only on the most recent $W$ days;
$W$ is a hyperparameter with a knee under drift. (Ch.~\ref{ch:online})

\item[rsig block] Our engineered lag-difference features: multi-scale
momentum (SMA4 minus SMA20; SMA20 minus SMA120) and venue spread, all
from yesterday's responder path. (Ch.~\ref{ch:featurepatterns})

\item[Seed noise] Score variation across random seeds at fixed data and
config ($\pm 0.0003$--$0.0008$ here); the unit in which all claimed
effects must be measured. (Ch.~\ref{ch:problem})

\item[Shift test] Leak detector: recompute features shifted one step
forward; genuine signal degrades gracefully, leaked signal collapses.
(Ch.~\ref{ch:leakage})

\item[Signature (path signature)] The sequence of iterated integrals of
a path; a universal feature map for sequential data. Elegant; redundant
next to recurrent state in this campaign. (Ch.~\ref{ch:signatures})

\item[SMA / EWMA] Simple moving average (uniform window $w$; group delay
$(w{-}1)/2$) / exponentially weighted moving average (span
$\approx 2/\alpha - 1$). Both low-pass filters. (Ch.~\ref{ch:smoothing})

\item[Stability gate] The requirement that any structure (feature
ranking, clustering, selection) reproduce across temporally distant
windows before being built upon. (Ch.~\ref{ch:featforensics})

\item[Static vs.\ online score] Evaluation of a frozen model vs.\ the
same model with the adaptation procedure running; report both, their
gap is diagnostic. (Ch.~\ref{ch:validation}, \ref{ch:online})

\item[Temporal bagging] Ensemble members trained on different
(overlapping) date subsets of one precomputed pool; cheap decorrelation.
(Ch.~\ref{ch:ensembling})

\item[Trailing feature] Computed strictly from data before $t$
(rolling statistics, lags). The default-legal feature type.
(Ch.~\ref{ch:leakage})

\item[Variate token (inverted attention)] Treating each feature (rather
than each timestep) as an attention token, so attention models
cross-feature structure; made causal by applying it within single
timesteps. (Ch.~\ref{ch:attention})

\item[Walk-forward] Evaluation that proceeds through validation days in
calendar order, updating whatever the deployment protocol would update.
(Ch.~\ref{ch:validation})

\item[Weighted zero-mean $R^2$] This competition's metric:
$1 - \sum w(y - \hat y)^2 / \sum w y^2$. Zero point = predicting zeros;
rewards shrinkage; punishes bias quadratically. (Ch.~\ref{ch:metric})

\end{description}
