\chapter{Feature Selection and Discovery}
\label{ch:selection}

More features are not better. Every column you feed a model is variance you
must pay for out of a signal budget of one percent; redundant columns
dilute tree ensembles and slow everything; unstable columns are leakage's
quieter cousin, borrowing score from relationships that will not survive to
test time. This chapter presents the selection pipeline that won the DRW
crypto competition, our port of it, and the places where our forensic
toolkit had to correct it.

\section{The DRW pipeline, annotated}

The winner faced $\sim$890 anonymous features and distilled them roughly as
follows --- annotations ours:

\subsection{Step 1: correlation clustering to medoids}

Compute the feature correlation matrix; define distance $1 - |\rho|$;
cluster (he solved a combinatorial medoid problem; complete-linkage
hierarchical clustering cut at distance $0.4$, i.e.\ $|\rho| \ge 0.6$, is
the practical equivalent); keep one representative per cluster --- the
\textbf{medoid}, the member with the highest total $|\rho|$ to its
clustermates, ``the most expensive point to replicate.'' 890 features
$\to$ $\sim$60.

\emph{Annotation:} the right first move on a raw, redundant dump ---
correlated duplicates split importance in trees, waste attention in
networks, and multiply the selection search space. But see
\S\ref{sec:selpitfalls} for what happened when we applied it to an
already-curated pool.

\subsection{Step 2: near-zero-correlation cut}

Drop features with $|\corr(x, y)| \le 10^{-4}$. $\sim$60 $\to$ $\sim$40.
\emph{Annotation:} a threshold this tiny only removes the truly inert; it
is deliberately conservative because univariate correlation is a bad judge
of value (Chapter~\ref{ch:featforensics}) --- weak-univariate features can
be strong in combination.

\subsection{Step 3: SHAP-consistency across folds}

Fit a gradient-boosted model per CV fold (his: six purged bimonthly
groups); compute mean-$|$SHAP$|$ importance per fold; keep features in the
top-20 of \emph{every} fold; union across model variants. $\sim$40 $\to$
$\sim$30.

\emph{Annotation:} the load-bearing idea. Importance in \emph{one} fold is
importance in one regime --- possibly noise, possibly a relationship that
died in 2024. Consistency across temporally distinct folds is a
\emph{stability filter}, the same epistemology as our two-window atlas
gate (\S\ref{sec:atlas}), applied to importance instead of timing. Our
port keeps features in the top-20 of $\ge 3$ of 4 contiguous folds.

\subsection{Step 4: symbolic recombination}

Generate second- and third-order combinations of the survivors
($X + Y$, $X - Y$, $XY$, $X/Y$, $\min$, $\max$, $XYZ$, $(X+Y)/Z$, \dots),
screen by target correlation, re-run step 3 on the enlarged pool.

\subsection{Step 5: recycling}

Re-add previously dropped features \emph{one at a time} and keep any that
improve validation. \emph{Annotation:} this exists because features
\textbf{lose predictive power over time} --- the winner observed his older
feature set degrading on newer data and re-audited every discard.
Selection is not a one-shot decision; under drift it has a maintenance
schedule.

\subsection{Step 6: autoencoder features}

Feed the final set through a small autoencoder; append the bottleneck
activations ($\sim$8 dims) as features. His single biggest reported boost.
\emph{Annotation:} an AE bottleneck is a learned nonlinear factor model of
the feature set; its coordinates expose ``common state of the market'' in
a form trees can split on. It is also a black box inside your pipeline ---
his own caveat was that pipeline robustness matters most exactly here.

\section{Our port, and the two measured corrections}
\label{sec:selpitfalls}

Running the pipeline on our engineered pool
(\code{scripts/drw\_feature\_lab.py}; 134 columns in, same fixed harness as
Chapter~\ref{ch:featurepatterns}) reproduced the wins and exposed two
boundary conditions:

\begin{center}
\begin{tabular}{@{}lrr@{}}
\toprule
variant & features & $\Rw$ \\
\midrule
all 134 columns (baseline) & 134 & $+0.00591$ \\
medoids only & 87 & $+0.00576$ \\
medoids + combos & 127 & $+0.00586$ \\
all + combos (importance-seeded) & 174 & $+0.00630$ \\
\textbf{all + combos (atlas-seeded)} & 184 & $\bm{+0.00643}$ \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Correction 1: pruning a curated pool costs signal.} Medoid
reduction, the step that made the DRW winner's raw dump tractable,
\emph{lost} $0.00015$ here: our 134 columns were already engineered with
intent, and their redundancy was partly informative (a feature and its
market average are correlated \emph{and} both useful). Diagnosis: cluster
pruning is for raw dumps; for curated pools, keep everything and let the
search \emph{add}.

\textbf{Correction 2: importance-based seeding misses interaction-only
value.} The lag-responder columns never survive SHAP-consistency (as
levels they are worthless --- Chapter~\ref{ch:featurepatterns}), so the
symbolic stage never got the chance to difference them into the rsig
momentum block we know is worth $+0.0003$. And the correlation screen
inside step 4 has the same blind spot: a combination whose parents have
zero univariate correlation cannot beat its parents by a metric on which
both score zero-ish in a way that surfaces reliably. The fix that
measured: \textbf{seed the candidate pool from forensics} --- atlas lead
features, reversal block, lag columns --- alongside the importance
survivors ($+0.00039 \to +0.00052$). The general form: importance tells
you what the model already uses; \emph{forensics tells you what it
should}.

\begin{warnbox}[: selection is model selection, done many times]
Every screening decision is a model comparison on finite data, and a
pipeline that makes hundreds of such comparisons on one validation signal
is a machine for manufacturing selection bias
(Chapter~\ref{ch:leakage}, experiment hygiene). The DRW pipeline's
defenses --- fold-consistency, train-side screening with tail-side
validation touched once per stage, parent-beating requirements,
redundancy caps --- are not decoration; they are what keeps a
feature-generation loop from becoming an overfitting loop. Port the
defenses, not just the search.
\end{warnbox}

\section{A synthesis: the selection stack we recommend}

For a fresh competition, in order:

\begin{enumerate}
\item \textbf{Forensic atlas first} (Chapter~\ref{ch:featforensics}):
  timing taxonomy, stability gate. Cost: an afternoon. This is your prior.
\item \textbf{Raw dump?} Medoid-prune at $|\rho| \ge 0.6$.
  \textbf{Curated pool?} Skip pruning.
\item \textbf{Fold-consistent importance} (4--6 contiguous folds, top-$K$
  intersection-ish) to find the model-visible core.
\item \textbf{Symbolic search} over core $\cup$ forensic seeds; screen on
  train, redundancy-cap, validate each stage once on the tail.
\item \textbf{Ablate per model family} --- feature value is a
  (feature, model) property.
\item \textbf{Schedule recycling}: re-audit discards when the validation
  window rolls forward. Under drift, selection decays like everything
  else.
\item Optional: \textbf{AE bottleneck} over the final set --- high
  expected value by DRW's evidence, black-box risk priced in.
\end{enumerate}

\begin{jsbox}[: what the discovered combos looked like]
The atlas-seeded search's accepted list is itself a forensic document.
Fifteen of fifty combos paired a \emph{reversal} feature with a
\emph{market-wide} factor --- e.g.\
$\min(\feat{07}_{\text{mkt avg}},\ \feat{45})$ --- that is, ``reversal
signal, gated by market state.'' Nobody hypothesized that structure; the
search found it, the tail validated it, and it immediately suggests the
next experiments (conditional reversal models; market-state interaction
terms in the nets). A good selection pipeline is also a discovery
instrument.
\end{jsbox}
