\chapter[Path Signatures]{The Model Zoo IV: Path Signatures --- a Beautiful Negative Result}
\label{ch:signatures}

This chapter is a full post-mortem of an idea that was mathematically
elegant, correctly implemented, individually competitive --- and added
nothing. We include it at chapter length deliberately: the failure mode it
illustrates (\emph{redundancy with an existing model family}) is one of the
genre's most expensive lessons, and the machinery itself remains worth
knowing for problems shaped differently than this one.

\section{The mathematics, briefly}

The \textbf{signature} of a path $X: [0, T] \to \mathbb R^d$ is the
sequence of iterated integrals
\[
S^{(i_1, \dots, i_k)}(X) \;=\;
\idotsint\limits_{0 < t_1 < \cdots < t_k < T}
dX^{(i_1)}_{t_1} \cdots\, dX^{(i_k)}_{t_k},
\qquad k = 1, 2, \dots
\]
collected over all index words $(i_1,\dots,i_k)$. Level 1 gives net
increments; level 2 encodes pairwise areas (which coordinate led which);
higher levels encode progressively finer order-of-events geometry. Two
properties make signatures theoretically seductive for us:

\begin{itemize}
\item \textbf{Universality}: continuous functionals of a path are
  approximated by \emph{linear} functionals of its signature --- a
  canonical, model-free feature map for sequences.
\item \textbf{Compression of order information}: exactly the ``what
  happened, in what order, how fast'' content that intraday context
  consists of.
\end{itemize}

Practical refinements we implemented (in \code{janestreet.theory
.signatures}): truncation at depth 2--3 over $\le 8$ selected channels
(the level-$k$ term count grows as $d^k$); the \textbf{log-signature} in a
Lyndon-word basis (removes algebraic redundancy among signature terms);
and a \textbf{Volterra/Hurst reweighting} --- multiplying increments by
$(t - s)^{H - 1/2}$ before integration --- so that recent path segments
dominate, mimicking rough-kernel memory with $H \approx 0.1$.

\section{What we built and what it scored}

The signature block computed, per timestep, the (log-)signature of the
trailing window (16--64 steps) of selected feature channels, and fed it as
extra inputs to either an MLP (\code{mlp\_sig}) or a causal transformer
(\code{sig\_transformer}). Channel selection mattered and taught its own
lesson:

\begin{itemize}
\item Selecting channels by \emph{univariate target correlation} made
  things \textbf{worse} than an arbitrary default ($+0.0044$ vs
  $+0.0049$): the signature's value lives in cross-channel monomials of
  \emph{path-rich} series, not in marginally predictive ones ---
  a screen aimed at the wrong property.
\item Selecting by an \textbf{autoregressive criterion} --- ridge-fit each
  feature's own lags against the target, rank by that $R^2$ --- found
  channels whose \emph{paths} carry information: $+0.0059$ static, the
  best signature variant. (Echo of Chapter~\ref{ch:selection}: the screen
  must match the mechanism.)
\item The refit-learning-rate cliff bit hard here first: at the RNNs'
  refit lr the sig-transformer \emph{lost} $0.0015$ online, and at
  $10\times$ smaller it gained $+0.0004$ --- the diagnosis that became
  Chapter~\ref{ch:online}'s family-specific-cliff law. Final best:
  $+0.0062$ online.
\end{itemize}

$+0.0062$ is a respectable single model --- above XGB, below the RNNs.
Then came the only test that matters for a marginal family:

\section{The ensemble verdict}

Added to the standing three-stream ensemble (xgb + two online RNNs,
$+0.0104$):

\begin{center}
\begin{tabular}{@{}lr@{}}
\toprule
ensemble & $\Rw$ \\
\midrule
3-way (incumbent) & $+0.0104$ \\
+ sig-transformer stream (4-way average) & $+0.0101$ \\
optimal-weight blend, 4 streams & assigns signature weight $\approx 0$ \\
\bottomrule
\end{tabular}
\end{center}

The signature stream's predictions correlated with the RNN streams tightly
enough that averaging it in \emph{diluted} the blend. The information the
signature extracts from an intraday window --- increments, leads/lags,
order-of-events geometry --- is evidently a subset of what a gated RNN's
hidden state already carries by that point in the day. The mathematics is
not wrong; it is \emph{redundant given the incumbents}.

\begin{keybox}[: elegance is not additivity]
The question for any new family is never ``is it good?'' but ``is it good
\emph{and decorrelated from what I have}?''
(Chapter~\ref{ch:ensembling}'s admission rule). A solo score can be earned
by re-deriving information the ensemble already owns; only decorrelated
error buys anything. Probe correlation \emph{before} polishing: two seeds,
small config, correlate the validation prediction stream against each
incumbent --- an afternoon that would have saved us several weeks.
\end{keybox}

\section{Where signatures still belong}

The machinery survives in the repository (\code{janestreet.theory}), and
the honest map of where it should pay:

\begin{itemize}
\item \textbf{No recurrent incumbent}: pipelines built on trees/MLPs only
  (many production settings) --- signature features are the cheapest way
  to hand path geometry to a stateless model; our \code{mlp\_sig} at
  $+0.0066$ beat XGB with zero recurrence.
\item \textbf{Short, irregular, or event-driven sequences} where RNN state
  has nothing to accumulate but path geometry still discriminates
  (medical time series, sensor bursts).
\item \textbf{As a screen}: the AR-based channel selector we built for
  signatures outlived them --- it is a general ``whose \emph{path} is
  informative?'' ranking, complementary to the atlas.
\end{itemize}

\section{Post-mortem: the process failure, not the math failure}

The reconstruction of how a wrong bet stayed alive for weeks, because the
pattern is general:

\begin{enumerate}
\item \textbf{Solo-score progress masked ensemble irrelevance.} Each
  iteration improved the signature model against itself; nobody re-ran the
  correlation probe until late.
\item \textbf{A confounded failure was mis-attributed.} The online-refit
  regression looked ``architectural'' for two runs before the lr sweep
  revealed a tuning cliff --- during which time the family's stock was
  wrongly discounted, muddying decisions.
\item \textbf{Sunk elegance.} The mathematics was a pleasure, and pleasure
  is a bias. The eventual kill decision was made by one number (blend
  weight $\approx 0$) that had been computable from week one.
\end{enumerate}

The repaired process --- decorrelation probe first, family-specific lr
sweep before any online verdict, kill criteria written down in advance ---
is codified in Chapters~\ref{ch:online}, \ref{ch:ensembling}, and
\ref{ch:diagnostics}. That is what a negative result is \emph{for}.
