\chapter{Ensembling}
\label{ch:ensembling}

Chapter~\ref{ch:problem} argued that at low SNR, estimation variance is the
dominant error term and averaging is the dominant cure. This chapter turns
that argument into procedure: what to average, how to weight, how to
manufacture diversity cheaply, and the two seductive upgrades (fitted
weights, stacking) that measured \emph{worse} than doing nothing clever.

\section{The arithmetic of averaging}

For prediction streams $\hat y^{(1)}, \dots, \hat y^{(m)}$ with equal
individual error variance $\sigma^2$ and average pairwise error correlation
$\bar\rho$, the simple average's error variance is
\[
\Var\!\Big(\tfrac{1}{m}\sum_i \varepsilon^{(i)}\Big)
= \sigma^2\Big(\frac{1-\bar\rho}{m} + \bar\rho\Big)
\;\xrightarrow{m \to \infty}\; \sigma^2 \bar\rho .
\]
Two readings, both of which drove decisions:

\begin{itemize}
\item The benefit saturates fast in $m$ but is \emph{unbounded} in
  $(1-\bar\rho)$: after three or four members, adding more of the same
  helps little, while adding something \emph{different} keeps paying.
  Diversity is the currency; member count is not.
\item A weaker model with decorrelated errors can beat a stronger clone.
  Our XGB stream scores $+0.0068$ against the LSTM's $+0.0092$, yet
  removing XGB from the blend costs more than removing the second RNN ---
  tree errors and neural errors disagree in structurally different
  places.
\end{itemize}

\begin{jsbox}[: the headline blend]
Three streams --- xgb static ($+0.0068$), gru\_modelr online ($+0.0089$),
lstm\_modelr online ($+0.0092$) --- simple average: $\bm{+0.0104}$.
The blend beats its best member by $+0.0012$, which is more than almost
any single modeling idea in this book earned. Cost of the technique:
one line of numpy.
\end{jsbox}

\section{Simple average versus fitted weights: a measured upset}

Optimal linear blending --- fit weights on held-out data --- dominates the
simple average \emph{in expectation with enough data}. At this SNR you do
not have enough data. Our measurement: weights fit on one half of the
validation tail, evaluated on the other half, \textbf{lost to the simple
average every time}, and the fitted weights themselves told the story ---
they kept collapsing into corner solutions (all weight on one stream, zero
on whole families), the classic signature of an optimizer fitting noise.
Weight estimation is itself a low-SNR regression with $m$ parameters and
one noisy signal; with 3--6 streams the estimation error exceeds the gain
over uniform weights.

Non-linear stacking (an XGB meta-model over the streams) did worse still,
for the same reason squared: more meta-capacity, same meta-noise.

\begin{keybox}[: the ensemble complexity ladder]
Climb only while a held-out set (that the weights never touched) says the
step pays: (1) simple average $\to$ (2) sign-constrained coarse weights
(e.g.\ family weights in $\{0.5, 1, 1.5\}$) $\to$ (3) fitted convex
weights $\to$ (4) stacking. In this campaign the ladder ended at rung 1.
The DRW winner's ensemble was similarly primitive: NN protagonist, XGB
seasoning. Public post-mortems of this genre are a graveyard of rung-4
blends that lost to rung 1 on private data.
\end{keybox}

\section{Manufacturing diversity}

Ranked by measured or observed cost-effectiveness in this campaign:

\begin{enumerate}
\item \textbf{Family diversity} (trees + RNN): the largest decorrelation
  per unit effort --- different function class, different failure
  geometry. This is why XGB is never dropped
  (Chapter~\ref{ch:treesmlp}).
\item \textbf{Seed diversity}: same architecture, different seeds. Cheap,
  reliable, small: the reference 8th-place solution averaged
  \emph{sixteen} seeds of one architecture. At $\pm 0.0005$ seed spread,
  averaging 4--16 seeds recovers a meaningful fraction of it.
\item \textbf{Temporal bagging}: members trained on different
  (overlapping) date subsets --- distinct regimes produce distinct
  errors, and the memmap manifest
  (Chapter~\ref{ch:preprocessing}) makes it one config line
  (\code{resample: subsample, frac 0.6}). Our infrastructure treats
  ``bag = list of (model, seed, date-resample) members'' as the unit of
  experiment.
\item \textbf{Adaptation diversity}: static and online variants of the
  \emph{same} model decorrelate on exactly the days that drift ---
  including both is nearly free since one is a byproduct of evaluating
  the other.
\item \textbf{Architecture diversity beyond family}: the expensive
  aisle, to be shopped with the admission rule below --- our signature
  and vanilla-transformer streams both failed it after substantial
  investment (Chapter~\ref{ch:signatures}).
\end{enumerate}

\section{The admission rule}

Codifying the campaign's most-regretted omission into procedure: a
candidate stream earns ensemble admission by showing, on cheap probes
(2 seeds, small config), that it is

\begin{enumerate}
\item \textbf{non-negative solo} (it clears zero robustly), and
\item \textbf{decorrelated}: prediction-stream correlation with every
  incumbent below a threshold we set at $\sim 0.9$, and --- the sharper
  test --- the incumbent blend \emph{with} the candidate beats the blend
  without it on held-out days.
\end{enumerate}

Run the probe \emph{before} polishing the candidate
(Chapter~\ref{ch:signatures}'s post-mortem: solo-score progress is a
treadmill if the stream is redundant). Notice the rule's corollary: a
mediocre-looking model can be admitted (XGB) while an impressive one is
rejected (sig-transformer, $+0.0062$ solo, blend weight zero). The
ensemble does not care about résumés.

\section{Ensembling adaptive models}

Two operational notes specific to this genre's online dimension
(Chapter~\ref{ch:online}):

\begin{itemize}
\item \textbf{Blend after adaptation, not before}: each member refits
  itself daily, then predictions are averaged. Averaging \emph{weights}
  across members (as opposed to predictions) is a different, riskier
  operation --- only valid within identical architectures, and unneeded
  here.
\item \textbf{Diversity drifts too.} Streams that decorrelate today may
  converge as they all adapt toward the same recent regime. If the test
  period is long, re-check pairwise correlations on a rolling basis ---
  in our walks the effect was visible but small; over months it might
  not be.
\end{itemize}

\section{Checklist}

\begin{enumerate}
\item Build the simple average of your best-per-family models before any
  further modeling; it is the new baseline.
\item Admit streams by the two-part rule (non-negative solo;
  decorrelation with incumbents), probed cheaply, before polishing.
\item Manufacture diversity in cost order: families, seeds, temporal
  bags, adaptation variants --- architectures last.
\item Climb the weighting ladder only while untouched held-out data says
  each rung pays; expect to stop at rung 1.
\item Keep the blend's member list short and auditable; log per-member
  and blend scores per day so a degrading member is visible.
\end{enumerate}
