\chapter[The Graveyard of Negative Results]{The Graveyard: Negative Results, Properly Buried}
\label{ch:negative}

A negative result that is written down with its evidence is an asset: it
prunes the search space forever. One that is merely remembered gets
re-litigated every time enthusiasm returns. This chapter is our graveyard
--- each entry with the tombstone (what died), the autopsy (why), and the
epitaph (the general rule it purchased). Scores are weighted $R^2$ on the
standard harness.

\section{Training on more history}

\textbf{Tombstone.} Curriculum training across 800 days of history:
$+0.0077$, \emph{versus} $+0.0089$ for the same architecture on the recent
280 days. Old-data models' \emph{static} scores on the recent tail were
frequently negative. The window sweep (50/100/200/400 days $\to$
$+0.0032/+0.0054/+0.0059/+0.0071$ online) rises toward a knee and the
full-history point falls off it.

\textbf{Autopsy.} Drift (\S\ref{sec:drift}): old regimes teach
feature--target relationships that no longer hold, and at low SNR the
model cannot tell stale structure from current structure --- extra rows
add interference, not information.

\textbf{Epitaph.} \emph{More data is not a default good under drift; the
training window is a hyperparameter with a knee. Sweep it.}

\section{Symbol identity as a feature}

\textbf{Tombstone.} One-hot symbol id into XGB: $-0.00002$.

\textbf{Autopsy.} Per-symbol rolling normalization
(Chapter~\ref{ch:preprocessing}) had already removed the per-symbol level
--- identity's main content. Thirty-nine dummy columns bought tree splits
and nothing else. The salvageable remainder was the \emph{stable} coarse
grouping: feature-profile clusters (+0.00022).

\textbf{Epitaph.} \emph{Before adding a feature, ask what your
normalization already removed. Identity features prosper only where
identity-linked structure survives preprocessing --- and is stable.}

\section{Symbol clustering by responder co-movement}

\textbf{Tombstone.} Clusters computed on one window looked excellent
(intra/inter co-movement ratio $\sim$18$\times$); recomputed on a later
window, agreement with the earlier clustering was ARI $= -0.07$ ---
chance. The planned per-cluster specialist models were cancelled without
training one.

\textbf{Autopsy.} Co-movement structure in this panel is
regime-transient. A routing scheme built on it routes tomorrow's symbols
by yesterday's dead map.

\textbf{Epitaph.} \emph{Any structure you intend to build on must first
pass a cross-window stability gate (\S\ref{sec:atlas}). Ten minutes of
checking beats weeks of specialist models. Compare the atlas: same gate,
Spearman $+0.97$ --- build on that instead.}

\section{Raw lagged responders as flat features}

\textbf{Tombstone.} Nine previous-day responder columns into XGB:
$-0.00003$.

\textbf{Autopsy.} As \emph{levels} at matched clock time, yesterday's
moving averages carry almost no linear information about today's value
(the construction: MAs of near-white innovations). Their content is
relational --- differences across windows and venues ($+0.0003$ as the
rsig block) or sequential (as inputs to a recurrent/endogenous stream).

\textbf{Epitaph.} \emph{A dead feature is often a live feature wearing
the wrong parameterization. Consult the construction before discarding
the information} (Chapter~\ref{ch:featurepatterns}).

\section{Bigger models}

\textbf{Tombstone.} LSTM width doubled (0.9M $\to$ 1.6M parameters):
score change within seed noise. Depth additions: same.

\textbf{Autopsy.} The binding constraints are the signal budget and
estimation variance, not expressiveness
(Chapter~\ref{ch:problem}). Capacity beyond ``enough'' buys variance.

\textbf{Epitaph.} \emph{At low SNR, capacity sweeps are flat and
procedure sweeps are not (Chapter~\ref{ch:online}). Spend tuning hours
where the curvature is.}

\section{Signature models as ensemble members}

\textbf{Tombstone.} Best signature variant $+0.0062$ solo; ensemble
weight $\approx 0$; 4-way average \emph{below} the 3-way ($+0.0101$ vs
$+0.0104$). Full post-mortem: Chapter~\ref{ch:signatures}.

\textbf{Epitaph.} \emph{Admission is decorrelation, not solo score. Probe
correlation before polishing.}

\section{Non-linear stacking and fitted blend weights}

\textbf{Tombstone.} Fitted convex weights: corner solutions, lost to the
simple average on held-out halves. XGB meta-stacker: worse.

\textbf{Autopsy.} Weight fitting is itself low-SNR estimation; with a
handful of streams the estimation error exceeds the edge over uniform
(Chapter~\ref{ch:ensembling}).

\textbf{Epitaph.} \emph{The ensemble ladder is climbed only while
untouched data pays you to climb; expect rung one.}

\section{Univariate correlation as a feature screen}

\textbf{Tombstone.} Signature channels ranked by $|\corr(x, y)|$
underperformed an arbitrary default ($+0.0044$ vs $+0.0049$); lag
columns, valuable in combination, score zero on the same screen and were
nearly lost to it (Chapter~\ref{ch:selection}).

\textbf{Autopsy.} The screen conflates nowcast with lead
(\S\ref{sec:atlas}) and is blind to interaction-only value.

\textbf{Epitaph.} \emph{Match the screen to the mechanism you are
selecting for: timing screens for timing (atlas alpha bands), path
screens for path models (AR-based), tree-gain screens for interactions.}

\section{A borrowed refit learning rate}

\textbf{Tombstone.} ``Online refit breaks signature-transformers'' ---
believed for two runs; actually a $10\times$ lr mismatch; at
$3\times10^{-5}$ the regression became a gain
(Chapter~\ref{ch:online}).

\textbf{Epitaph.} \emph{No cross-family conclusion is valid while a
procedure hyperparameter is borrowed. Sweep first, conclude second.}

\section{What the graveyard is for}

Notice the pattern across tombstones: almost every death was caused by
one of four recurring pathogens --- \textbf{drift} (sections 1, 3),
\textbf{already-normalized-away information} (2, 4), \textbf{low-SNR
estimation variance} (5, 6, 7), and \textbf{confounded procedure
hyperparameters} (8, 9). The graveyard is not a list of things that are
false; it is the empirical fingerprint of this genre's four ways of
fooling you. When your next idea dies, check which pathogen got it ---
and when your next idea \emph{succeeds}, check the same list before
believing it.

Maintain your own graveyard file from day one
(\code{WHAT\_DIDNT\_WORK.md} in our repository). The discipline costs a
paragraph per burial. It pays every time a promising-looking corpse
knocks on the door wearing a new hat --- which, over a long campaign, is
weekly.
